Concurrent Programming with CXL

### 2020-05-30 book/paper.tex
Started working on semaphores

@@

So far we have looked at how to protect a single resource.  Sometimes
we have multiple but a limited amount of resources that require
protection.  That is, if there are only $n$ resources, no more than
$n$ processes can access them at a time.

>>>>

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A simple counter is a very simple example of
a data structure.  A more involved one would be access to a binary tree.
Adding a node to a binary tree or re-balancing a tree often requires
multiple operations.  Maintaining consistency is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.

@@

\begin{figure}
\begin{code}
\begin{verbatim}
def Semaphore(cnt):
    result = cnt;
;
def P(sema):
    let blocked = True:
        while blocked:
            atomic:
                if (^sema) > 0:
                    ^sema = (^sema) - 1;
                    blocked = False;
                ;
            ;
        ;
    ;
;
def V(sema):
    atomic:
        ^sema = (^sema) + 1;
    ;
;
\end{verbatim}
\end{code}
\caption{The \texttt{Semaphore} interface in the \texttt{synch} module.}
\label{fig:semaphore}
\end{figure}

\begin{figure}
\begin{code}
\verbatiminput{../PCsema.cxl}
\end{code}
\caption{Bounded Buffer implementation using semaphores.}
\label{fig:boundedbuffer}
\end{figure}

So far we have looked at how to protect a single resource.  Sometimes
we have multiple but a limited amount of resources that require
protection.  That is, if there are only $n$ resources, no more than
$n$ can be used at a time; if all are in use and a process come along
that needs one of the resources, it has to wait until another process
releases one of the resources.  Note that we cannot solve this problem
simply using a lock per resource---allocating a resource requires access
to all of them.

A good example is the so-called \emph{bounded buffer}.  It is essentially
a queue implemented using a circular buffer of a certain length and two pointers:
one where items are inserted and one from where items are extracted.  If the
buffer is full, processes that want to add more items have to wait.  In the usual
formulation, processes that want to extract an item when the buffer is empty
also have to wait.  It would be easy to build such a thing using a busy waiting
approach, but again, we would like to avoid busy waiting.

Introduced by the famous Dutch computer scientist Edsger Dijkstra,
a \emph{semaphore} is a primitive that fits the bill.  A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
resources available.  When allocating a resource, a process decrements the
counter using the \texttt{procure} or simply \texttt{P} operation.  The
\texttt{procure} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{vacate} or {V} operation.
Figure~\ref{fig:sema} shows the \texttt{synch} module implementation of
semaphores.
